\bf Visualising How Non-linear Dimension Reduction Warps Your Data

### 2024-03-06 paper.tex
Make the plan for the methodology

\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[enhanced, interior hidden, borderline west={3pt}{0pt}{shadecolor}, breakable, boxrule=0pt, sharp corners, frame hidden]}{\end{tcolorbox}}\fi

@@

In this paper, we present a novel method that aims to determine the most
effective way to represent high-D data by selecting the best method and
hyperparameter choice. Our approach involves dividing the data set into
two parts: a training set to construct the model and a test set to
generate predictive values and residuals. To implement our approach, we
first use a 2D embedding data set as the initial point. The output of
our algorithm is a tour that displays the model and original data in the
high-dimensional space. A flow chart of the proposed algorithm is shown
in Figure~\ref{fig-meth}. Our algorithm comprises two main phases: (1)
generate the model in the 2D space, and (2) generate the model in the
high-D space. These two phases are described in detail in this section
using UMAP 2D embedding of the S-curve data set, which has seven
dimensions, including four noise dimensions.

@@

\caption{\label{fig-meth}A flow diagram detailing the steps taken to
create the low-dimensional manifold in the high dimensional space. There
are two basic phases, one to generate the model in the 2D space, and
other to generate the model in the high-D space.}

@@

@@

\textbf{(a) Determine the number of bins along the x-axis} (\(b_1\))

@@

\begin{equation}\protect\hypertarget{eq-equation3}{}{
 \text{A} = \frac{\sqrt{3}}{2}h^2
}\label{eq-equation3}\end{equation}

To construct regular hexagons, \(A = 1\) (see Figure~\ref{fig-binsize})
use as the default setting. Then, the diameter (\(h\)) of the regular
hexagons is calculated (see Equation~\ref{eq-equation4}).

\begin{equation}\protect\hypertarget{eq-equation4}{}{
  \text{h} = \sqrt{\frac{2}{\sqrt{3}}A}
}\label{eq-equation4}\end{equation}

\citet{Carr2013} mentioned about the relationship between the diameter
(\(h\)) of regular hexagons and the height (\(y\)) of the plotting
region. According to our algorithm, the height (\(y\)) of the plotting
region is the the range of 2D embedding component 1 (\(r_1\)) (see
Equation~\ref{eq-equation6}).

\begin{equation}\protect\hypertarget{eq-equation6}{}{
h = \frac{r_1}{b_1}
}\label{eq-equation6}\end{equation}

After rearranging the Equation~\ref{eq-equation6} as
Equation~\ref{eq-equation5}, \(b_1\) is computed. The \(b_1\) value is
rounded up to the nearest whole number to have an integer value.

\begin{equation}\protect\hypertarget{eq-equation5}{}{
b_1 = \frac{r_1}{h}
}\label{eq-equation5}\end{equation}

\textbf{(b) Determine the shape parameter} (\(s\))

In this step, we determine the shape parameter (\(s\)) for the hexagonal
bins, which significantly influences their shape and arrangement within
the grid. The \(s\) in the hexagonal binning algorithm is defined as the
ratio of the height (\(y\)) to the width (\(x\)) of the plotting region
as defined in Equation~\ref{eq-equation1}.

\begin{equation}\protect\hypertarget{eq-equation1}{}{
s = \frac{y}{x}
}\label{eq-equation1}\end{equation}

The shape parameter (\(s\)) of our algorithm is calculated as the ratio
of the ranges of 2D embedding components, where \(r_1\) and \(r_2\)
represent the ranges of 2D embedding components 1 and component 2,
respectively (see Equation~\ref{eq-equation2}).

\begin{equation}\protect\hypertarget{eq-equation2}{}{
s = \frac{r_2}{r_1}
}\label{eq-equation2}\end{equation}

\textbf{(c) Determine the number of bins along the y-axis} (\(b_2\))

Next, the number of bins along the y-axis is computed based on the
number of bins along the x-axis (\(b_1\)) and the shape parameter
(\(s\)) (see Equation~\ref{eq-equation12}) \citep{Carr2013}.

\begin{equation}\protect\hypertarget{eq-equation12}{}{
b_2 = 2 * \left(\frac{(b_1 \times s)}{\sqrt(3)} + 1.5001 \right)
}\label{eq-equation12}\end{equation}

\textbf{(d) Determine the total number of bins} (\(b\))

The total number of bins is determined by multiplying the number of bins
along the x-axis (\(b_1\)) with the number of bins along the y-axis
(\(b_2\)) (see Equation~\ref{eq-equation13}).

\begin{equation}\protect\hypertarget{eq-equation13}{}{
b = b_1 \times b_2
}\label{eq-equation13}\end{equation}

\textbf{Step 2: Obtain bin centroids}

@@

Expanding upon the information regarding hexagonal binning discussed in
Step 1 of Section~\ref{sec-construct2d}, the primary objective in this
step is to determine the 2D embedding points associated with each
hexagon. As the initial process of hexagonal binning, the 2D points are
clustered into their respective hexagonal bins. By mapping this
information with the hexagonal bin centroids
(\(C_k^{(2)} \equiv (C_{ky_1}, C_{ky_2})\)) that obtained in Step 2 of
the 2D model building (see Section~\ref{sec-construct2d}), we can find
which 2D points are assigned to which data set (see
Figure~\ref{fig-wkhighD} (a)).

@@

Following the step 1, the main focus in this step to determine the
corresponding high dimensional points for each hexagon. Every 2D
embedding point serves as a projection of a data point belonging to high
dimensional space. By using this mapping between the high dimensional
data and their corresponding projections, the high dimensional points
allocated to each hexagons are determined (see video linked in
Figure~\ref{fig-wkhighD}).

@@

Having identified the high-dimensional points associated with hexagons,
the final step involves computing the mean within each hexagonal bin.
This implies calculating the average of the high-dimensional data points
located within each hex bin. These averaged high-dimensional data
points, denoted as \(C_k^{(p)} \equiv (C_{kx_1}, ..., C_{kx_p})\), serve
as the representative coordinates for the hex bin centroids within the
expansive high-dimensional space (see video linked in
Figure~\ref{fig-wkhighD}).

@@

Based on the insights gained in Step 3 of Section~\ref{sec-construct2d},
where connected edges in the 2D triangular mesh were identified, this
step involves lifting these connections into the high dimensions. Having
the mappings of all 2D triangular mesh points into high dimensions, the
points connected in 2D are also connected in high dimensional space (see
video linked in Figure~\ref{fig-wkhighD}).

{\centering \includegraphics{paper_files/figure-pdf/fig-wkhighD-1.pdf}

\caption{\label{fig-wkhighD}How the 2D model lift into high dimensions?
(a) visualize the points and the hexagonal bin centroids related 2nd and
15th hexagons, (b) visualization of the edge connected the 2nd and 15th
hexagons (colored in red) in the triangular mesh. A video of tour
animation is available at \url{https://youtu.be/ffWFVK1UTIc}.}

@@

The number of hexagonal bins in the hexagonal grid has a considerable
impact on the construction of the 2D model. This is because it is the
initial step in building the 2D model. The hexagonal grid with the
chosen total number of bins must be able to capture the structure of the
NLDR data. If the number of bins is too low, the model may not be able
to capture the structure of the NLDR data effectively (see
Figure~\ref{fig-binsize} (a)), while if there are too many bins, it may
result in over-fitting the individual points of the NLDR data (see
Figure~\ref{fig-binsize} (c)). Therefore, it is important to determine
an appropriate number of bins to build an effective model.

Furthermore, the total number of bins is determined by the number of
bins along the x-axis and y-axis (see Equation~\ref{eq-equation13}). To
calculate the effective total number of bins, the minimum and
approximate maximum number of bins along the x-axis are used as
candidate values. The minimum number of bins along the x-axis is \(1\).
The maximum number of bins along the x-axis can be estimated by taking
the square root of the number of NLDR data points. To investigate the
impact of the number of bins on MSE, the analysis is focused on the MSE
for different total numbers of bins. This range encompasses the total
number of bins calculated for the minimum and maximum values of the
number of bins along the x-axis.

Typically, the MSE plot will have a steep slope at the beginning,
indicating that a smaller number of bins causes a larger amount of
error. Then, the slope will gradually decline or level off, indicating
that a higher number of bins generates a smaller error. This MSE plot is
useful in determining the effective number of bins required to construct
the 2D model. Figure~\ref{fig-diagnosticpltScurve} (b) has created a
graph that shows how the Mean Squared Error (MSE) changes depending on
the number of bins assigned in the hexagonal grid for various NLDR
techniques applied to the S-curve data set. As per the rule of thumb
mentioned earlier, the effective number of bins for each NLDR method is
shown in the \textbf{?@tbl-msebinsnldr}. Furthermore,
\textbf{?@fig-modelScurve} shows the 2D models constructed for each NLDR
technique with the selected total number of bins and how the models look
in high-D with original data.

\begin{figure}[H]

{\centering \includegraphics{paper_files/figure-pdf/fig-binsize-1.pdf}

\caption{\label{fig-binsize}Hexbin plots from different number of bins
for the \textbf{UMAP} projections of \textbf{S-curve} training data: (a)
b = 32 (4, 8), s = 1.643542, (b) b = 264 (12, 22), s = 1.643542, and (c)
b = 840 (21, 40), s = 1.643542. The hexbins are colored based on the
density of points, with darker colors indicating higher point density
and yellow color representing lower point density within each bin. What
is the number of bins that would be effective in representing
low-dimensional data?}

{\centering \includegraphics{paper_files/figure-pdf/fig-diagnosticpltScurve-1.pdf}

\caption{\label{fig-diagnosticpltScurve}Goodness of fit statistics from
UMAP applied to training S-curve dataset. What is the effective number
of bins in each NLDR technique to create a 2D model? The MSE plot have a
steep slope at the beginning, indicating that a smaller number of bins
causes a larger amount of error. Then, the slope gradually declines or
level off, indicating that a higher number of bins generates a smaller
error. Using the elbow method, when the total number of bins is set to
144, the slope of the Mean Squared Error (MSE) plot experiences a sudden
and noticeable change, resembling an elbow-like shape. This point
indicates that adding less bins does not enough to capture the data
structure.}

@@

The benchmark value for removing low-density hexagons has a range of
\(0\) and \(1\). When analyzing how these benchmark values affect model
performance, it's important to observe the change in Mean Squared Error
(MSE) as the benchmark value increases. The MSE exhibits a gradual
increase as the benchmark value progresses from \(0\) to \(1\).
Assessing this rate of increase is crucial. If the increment is not
substantial, the decision might lean towards retaining low-density
hexagons. According to Figure~\ref{fig-diagnosticpltScurvelwd}, the
change in MSE for NLDR techniques is relatively small across different
benchmark values, except for the PHATE method.

Consequently, it is not necessary to remove low-density hexagons when
constructing the 2D model for the S-curve dataset in NLDR techniques,
except for PHATE. While there may be some fluctuations in MSE between
benchmark values of 0 to 1, the absence of significant changes suggests
that maintaining low-density hexagons is preferable. However, to choose
the best selection, it's recommended to examine the benchmark values
around the first local minimum in the MSE curve. However, the MSE for
PHATE fluctuates when benchmark values range between 0 and 1. Therefore,
it is necessary to explore the benchmark values surrounding the first
local minimum to determine an effective benchmark value. There is no
universal rule for selecting a process. As shown in
Figure~\ref{fig-diagnosticpltScurvelwd}, the MSE values for PHATE
fluctuate, and the first local minimum is at \(0\). Therefore, nearby
benchmark values such as \(0.01\), \(0.04\), and \(0.06\) were explored
to identify the best value for preserving the data structure. In this
case, a benchmark value of \(0.06\) was chosen. The resulting 2D model
is shown in \textbf{?@fig-modelScurve} (c), while the video link
provides a visual representation of how the model appears in high-D.

@@

\begin{figure}[H]

{\centering \includegraphics{paper_files/figure-pdf/fig-rmlowdenshex-1.pdf}

\caption{\label{fig-rmlowdenshex}Goodness of f}

{\centering \includegraphics{paper_files/figure-pdf/fig-diagnosticpltScurvelwd-1.pdf}

\caption{\label{fig-diagnosticpltScurvelwd}Goodness of fit statistics
from UMAP applied to training S-curve dataset with different benchmark
values to remove the low-density heaxgons. What is the effective
benchmark value to remove the low-density heaxgons? The MSE plot have a
steep slope at the beginning, indicating that a smaller benchmark causes
a small amount of error. Then, the slope gradually increases or level
up, indicating that a higher number of benchmark values generates a
higher error. Using the reverse elbow method, when the benchmark value
is set to 0.2, the slope of the Mean Squared Error (MSE) plot
experiences a sudden and noticeable change, resembling an elbow-like
shape. This point indicates that higher benchmark values remove the
necessary bins as well which lead to distruct the 2D structure.}

@@

Achieving a smooth representation in 2D space is crucial, and one factor
influencing this smoothness is the presence of long edges (see
Figure~\ref{fig-modelScurvermlgimp}). These long edges occur when a line
connects points that are distant in the triangular mesh, impacting the
overall smoothness of the mesh.

@@

{\centering \includegraphics{paper_files/figure-pdf/fig-modelScurverdistcount-1.pdf}

\caption{\label{fig-modelScurverdistcount}plots to identify benchmark
values}

{\centering \includegraphics{paper_files/figure-pdf/fig-modelScurvermlgimp-1.pdf}

\caption{\label{fig-modelScurvermlgimp}What does the 2D model,
constructed using UMAP (applied to the S-curve dataset), look like in
high dimensions with long edges? (\url{https://youtu.be/pSL_G_n7-fM})
(a) 2D model without removing the long edges, and (b) 2D model with long
edges coloured by red. The default benchmark value is used to identify
long edges.}

\hypertarget{hexagonal-grid-origin}{%
\subsubsection{Hexagonal grid origin}\label{hexagonal-grid-origin}}

According to \citet{Dan2023}, the hexagonal binning is done by
tessellating the \(xy\) plane over the set (range(\(x\)), range(\(y\)))
(see Figure~\ref{fig-scurveshifthexgridsexp} (b)). In that case, bin
centroids are defined as shown in
Figure~\ref{fig-scurveshifthexgridsexp} (a) with gray colour. Rather
than sticking to the typical hexagonal grid, introducing a meaningful
shift in both the x and y directions presents an opportunity for an
improved 2D model. Therefore, investigating this shift in the hexagonal
grid is an important parameter to consider.

As shown in Figure~\ref{fig-scurveshifthexgrids}, the shifting
influences the distribution of points and number of non-empty bins,
impacting the resulting 2D model. According to
Figure~\ref{fig-diagnosticpltScurvehexbins}, the 2D model with a total
of \(144\) bins applied to S-curve UMAP data does not require any
shifting because the lowest MSE occurs when no shift is introduced.

{\centering \includegraphics{paper_files/figure-pdf/fig-scurveshifthexgridsexp-1.pdf}

\caption{\label{fig-scurveshifthexgridsexp}(a) Visualization of bin
centroids and (b) hexagonal grid, showcasing the configuration before
(colored in gray) and after (colored in light orange) shifting the
starting point by an identical amount applied in both x and y
directions, with a shift value of \(0.537285\).}

{\centering \includegraphics{paper_files/figure-pdf/fig-scurveshifthexgrids-1.pdf}

\caption{\label{fig-scurveshifthexgrids}Hexbin plots shows the
distribution of points before and after shifting the starting point of
the hexagonal grid generated for UMAP applied to the training S-curve
dataset. (a) Hexbin plot before shifting, and (b) Hexbin plot after
shifting (identical shift applied in both x and y directions, with a
shift value of \(0.537285\)).}

{\centering \includegraphics{paper_files/figure-pdf/fig-diagnosticpltScurvehexbins-1.pdf}

\caption{\label{fig-diagnosticpltScurvehexbins}Goodness of fit
statistics from UMAP applied to training S-curve dataset with different
shift values. In this illustration, an identical shift is applied in
both the x and y directions. The Mean Squared Error (MSE) plot exhibits
a global minimum at \(0\), indicating that no additional shifting is
necessary for the current hexagonal grid to yield a well-fitted model.}

@@

In the 2D representations created by all NLDR techniques, as shown in
Figure~\ref{fig-nldervis5Gau}, except for PHATE, there are five distinct
clusters. In tSNE, these five clusters are closely located to each other
(see Figure~\ref{fig-nldervis5Gau} (a)). However, in PHATE, two clusters
are closely positioned, while the other three are more distant (see
Figure~\ref{fig-nldervis5Gau} (c)). In PaCMAP, one cluster is at the
center, and the remaining four are positioned in four different
directions (see Figure~\ref{fig-nldervis5Gau} (e)). In TriMAP, two
clusters are close, although not as close as in PHATE, and the other
three are well-separated (see Figure~\ref{fig-nldervis5Gau} (d)). In
UMAP, all clusters are arranged in a parallel manner, with three in one
line and the other two in a separate line (see
Figure~\ref{fig-nldervis5Gau} (b)).

Visualizing the models alongside the original high-D data provides
insights into how different techniques capture the underlying clustering
structure. The tSNE model exhibits five well-separated clusters,
effectively preserving both local and global structures (see video link
of Figure~\ref{fig-modelfiveGau} (a)). UMAP, on the other hand, also
presents five distinct clusters but with a more flattened surface
appearance (see video link of Figure~\ref{fig-modelfiveGau} (b)). The
PHATE model shows five separated clusters resembling triangles or
partial triangles but struggles to capture local structures (see video
link of Figure~\ref{fig-modelfiveGau} (c)). TriMAP and PaCMAP both
display five well-separated clusters with flat surfaces, each capturing
within-cluster variation to varying extents. Despite the various 2D
representations, all NLDR techniques preserve the global structure (see
video link of Figure~\ref{fig-modelfiveGau} (d), (e) respectively).
However, tSNE, effectively capture both local and global structures, as
indicated by lower AIC values in Figure~\ref{fig-diagnosticpltGau} (a).

{\centering \includegraphics[width=1\textwidth,height=\textheight]{paper_files/figure-pdf/fig-nldervis5Gau-1.pdf}

\caption{\label{fig-nldervis5Gau}2D layouts from different NLDR
techniques applied the same data: (a) tSNE (perplexity = 61), (b) UMAP
(n\_neighbors = 15), (c) PHATE (knn = 5), (d) TriMAP (n\_inliers = 5,
n\_outliers = 4, n\_random = 3), and (e) PaCMAP (n\_neighbors = 10, init
= random, MN\_ratio = 0.9, FP\_ratio = 2). Is there a best
representation of the original data or are they all providing equivalent
information?}

{\centering \includegraphics[width=1\textwidth,height=\textheight]{paper_files/figure-pdf/fig-modelfiveGau-1.pdf}

\caption{\label{fig-modelfiveGau}Is there a best model to represent the
original data in 2D space or are they all providing equivalent
information?, (a) Model in the 2D space with tSNE
(\url{https://youtu.be/x5VPB-wOLm4}), (b) Model in the 2D space with
UMAP (\url{https://youtu.be/Cm1dW6iLyHQ}), (c) Model in the 2D space
with PHATE (\url{https://youtu.be/HB0Y3ilz6sI}), (d) Model in the 2D
space with TriMAP (\url{https://youtu.be/iSVkXIrOrSA}), and (e) Model in
the 2D space with PaCMAP (\url{https://youtu.be/O2QBctsc4uM}).}

{\centering \includegraphics[width=1\textwidth,height=\textheight]{paper_files/figure-pdf/fig-diagnosticpltGau-1.pdf}

\caption{\label{fig-diagnosticpltGau}Goodness of fit statistics from
different NLDR techniques applied to training five spherical Gaussian
cluster dataset. What is the best NLDR technique to represent the
original data in 2D?}

@@

\includegraphics{paper_files/figure-pdf/unnamed-chunk-56-1.pdf}

@@

{\centering \includegraphics{paper_files/figure-pdf/fig-diagnosticpltPBMC2-1.pdf}

\caption{\label{fig-diagnosticpltPBMC2}Goodness of fit statistics from
UMAP applied to training S-curve dataset. What is the effective number
of bins in each NLDR technique to create a 2D model? The MSE plot have a
steep slope at the beginning, indicating that a smaller number of bins
causes a larger amount of error. Then, the slope gradually declines or
level off, indicating that a higher number of bins generates a smaller
error. Using the elbow method, when the total number of bins is set to
144, the slope of the Mean Squared Error (MSE) plot experiences a sudden
and noticeable change, resembling an elbow-like shape. This point
indicates that adding less bins does not enough to capture the data
structure.}

{\centering \includegraphics[width=1\textwidth,height=\textheight]{paper_files/figure-pdf/fig-nldervisPBMCUMAP2-1.pdf}

\caption{\label{fig-nldervisPBMCUMAP2}(a) 2D layout from UMAP
(n\_neighbors = 15, min\_dist = 0.99, metric = cosine) applied for the
PBMC3k dataset. Is this a best representation of the original data?, (b)
Model in the 2D space with UMAP (\textless\textgreater)}

@@

@@

\includegraphics{paper_files/figure-pdf/unnamed-chunk-71-1.pdf}

\hypertarget{harps-gto-sample-dataset}{%
\subsection{HARPS GTO sample dataset}\label{harps-gto-sample-dataset}}

\citet{Anders2018}

\begin{itemize}
\item
  In 2D, there are two well-separated clusters, but the clusters are
  really close
\item
  One cluster have a sparse data space within the cluster
\item
  There is a line in right corner of the structure that is not connected
  any other trimesh points. This line create a long edge in high-D
  space.
\item
  In high-D view, the clusters having some non-linearity patter, which
  is also not captured in tSNE. Therefore, this is not a good choice.
  Also, the two clusters are really near to each other and can't see any
  separation. Also continuity of the structure. Long edges are presence
  also.
\item
  The parameter is, in a sense, a guess about the number of close
  neighbors each point has.
\item
  Num bins along the x-axis = 43, shape parameter = 0.4466309
\end{itemize}

\caption{errors for different parameter combinations}
\centering
\begin{tabular}[t]{r|r}
\hline
\textbf{perplexity} & \textbf{error (x 100)}\\
\hline
\cellcolor{gray!6}{23} & \cellcolor{gray!6}{0.02}\\
\hline
91 & 0.04\\
\hline
\textcolor{red}{\textbf{\cellcolor{gray!6}{40}}} & \textcolor{red}{\textbf{\cellcolor{gray!6}{1.05}}}\\
\hline
\end{tabular}
\end{table}

{\centering \includegraphics{paper_files/figure-pdf/fig-diagnosticpltHARPS-1.pdf}

\caption{\label{fig-diagnosticpltHARPS}Goodness of fit statistics from
UMAP applied to training S-curve dataset. What is the effective number
of bins in each NLDR technique to create a 2D model? The MSE plot have a
steep slope at the beginning, indicating that a smaller number of bins
causes a larger amount of error. Then, the slope gradually declines or
level off, indicating that a higher number of bins generates a smaller
error. Using the elbow method, when the total number of bins is set to
144, the slope of the Mean Squared Error (MSE) plot experiences a sudden
and noticeable change, resembling an elbow-like shape. This point
indicates that adding less bins does not enough to capture the data
structure.}

{\centering \includegraphics{paper_files/figure-pdf/fig-nldervisharpstSNE-1.pdf}

>>>>

\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[enhanced, borderline west={3pt}{0pt}{shadecolor}, interior hidden, sharp corners, frame hidden, boxrule=0pt, breakable]}{\end{tcolorbox}}\fi

@@

In this paper, we introduce a novel method to determine the most
effective NLDR technique and the best hyperparameter choice that
provides the most useful representation of high-D data. Our approach
involves dividing the high-dimensional dataset into two parts: a
training set for constructing the model and a test set for generating
predictive values and residuals. Our algorithm takes a 2D embedding data
as the input and generate a tour that displays the high-dimensional
wireframe to overlay the data. The flow chart of the proposed algorithm
is shown in Figure~\ref{fig-meth}. The algorithm consists of two main
phases: (1) generating the model in the 2D space and (2) lifting the
model into high-D space. The main steps of the algorithm are described
in detail in this section using UMAP 2D embedding of the S-curve
dataset. This dataset has seven dimensions, including four noise
dimensions that were added to the original 3D data.

@@

\caption{\label{fig-meth}The flow diagram shows the main steps of our
algorithm. There are two basic phases, one to generate the model in the
2D space, and other to map the model into the high-D space.}

@@

\textbf{Step 1: Scaling NLDR data}

\textbf{Step 2: Hexagonating NLDR data}

\textbf{(a) Determine the number of bins along the x axis} (\(b_1\))

\textbf{(b) Determine the number of bins along the y axis} (\(b_2\))

\textbf{(c) Determine the total number of bins} (\(b\))

\textbf{(d) Assign NLDR data to hexagons}

\textbf{Step 3: Obtaining bin centroids or bin means}

\textbf{Step 4: Triangulating bin centroids or bin means}

@@

\textbf{(a) Determine the number of bins along the x axis} (\(b_1\))

@@

In the process of computing hexagonal bin configurations, the first step
involves determining the number of bins along the x-axis. This begins
with the initialization of parameters such as the hexagonal size (\(s\))
and the buffer along the x-axis. Subsequently, the range of the first 2D
embedding component (\(r_1\)) is calculated to ensure comprehensive
coverage of the data. This range is then adjusted by incorporating the
buffer amount, effectively expanding the boundary to accommodate
potential outliers or edge cases. The default buffer amount along the
x-axis is \(\sqrt{3} * s * 1.5\). Next, the horizontal spacing (\(h\))
between hexagons is computed based on the hexagon size. Finally, the
number of bins along the x-axis is computed based on the adjusted range
(\(r_1 + buffer_{x}\)) and the horizontal spacing (\(h\)).

\begin{equation}\protect\hypertarget{eq-equation3}{}{
 h = \sqrt{3} * s
}\label{eq-equation3}\end{equation}

\begin{equation}\protect\hypertarget{eq-equation4}{}{
 b_1 = \frac{r_1 + buffer_{x}}{h}
}\label{eq-equation4}\end{equation}

\textbf{(b) Determine the number of bins along the y-axis} (\(b_2\))

Next, the number of bins along the y-axis is computed based on the
number of bins along the x-axis (\(b_1\)) and the shape parameter
(\(s\)) (see Equation~\ref{eq-equation4}) \citep{Carr2013}.

\begin{equation}\protect\hypertarget{eq-equation5}{}{
  v = 1.5 * s
}\label{eq-equation5}\end{equation}

\begin{equation}\protect\hypertarget{eq-equation6}{}{
 b_2 = \frac{r_2 + buffer_{y}}{v}
}\label{eq-equation6}\end{equation}

\textbf{(c) Determine the total number of bins} (\(b\))

The total number of bins is determined by multiplying the number of bins
along the x-axis (\(b_1\)) with the number of bins along the y-axis
(\(b_2\)) (see Equation~\ref{eq-equation5}).

\begin{equation}\protect\hypertarget{eq-equation5}{}{
b = b_1 \times b_2
}\label{eq-equation5}\end{equation}

\textbf{Step 2: Obtain bin centroids or bin means}

@@

Expanding upon the information regarding hexagonal binning discussed in
Step 1 of Section~\ref{sec-construct2d}, the primary objective in this
step is to determine the 2D embedding points associated with each
hexagon. As the initial process of hexagonal binning, the 2D points are
clustered into their respective hexagonal bins. By mapping this
information with the hexagonal bin centroids
(\(C_k^{(2)} \equiv (C_{ky_1}, C_{ky_2})\)) that obtained in Step 2 of
the 2D model building (see Section~\ref{sec-construct2d}), we can find
which 2D points are assigned to which data set (see
\textbf{?@fig-wkhighD} (a)).

@@

Following the step 1, the main focus in this step to determine the
corresponding high dimensional points for each hexagon. Every 2D
embedding point serves as a projection of a data point belonging to high
dimensional space. By using this mapping between the high dimensional
data and their corresponding projections, the high dimensional points
allocated to each hexagons are determined (see video linked in
\textbf{?@fig-wkhighD}).

@@

Having identified the high-dimensional points associated with hexagons,
the final step involves computing the mean within each hexagonal bin.
This implies calculating the average of the high-dimensional data points
located within each hex bin. These averaged high-dimensional data
points, denoted as \(C_k^{(p)} \equiv (C_{kx_1}, ..., C_{kx_p})\), serve
as the representative coordinates for the hex bin centroids within the
expansive high-dimensional space (see video linked in
\textbf{?@fig-wkhighD}).

@@

Based on the insights gained in Step 3 of Section~\ref{sec-construct2d},
where connected edges in the 2D triangular mesh were identified, this
step involves lifting these connections into the high dimensions. Having
the mappings of all 2D triangular mesh points into high dimensions, the
points connected in 2D are also connected in high dimensional space (see
video linked in \textbf{?@fig-wkhighD}).

@@

The number of hexagonal bins in the hexagonal grid has a considerable
impact on the construction of the 2D model. This is because it is the
initial step in building the 2D model. The hexagonal grid with the
chosen total number of bins must be able to capture the structure of the
NLDR data. If the number of bins is too low, the model may not be able
to capture the structure of the NLDR data effectively (see
\textbf{?@fig-binsize} (a)), while if there are too many bins, it may
result in over-fitting the individual points of the NLDR data (see
\textbf{?@fig-binsize} (c)). Therefore, it is important to determine an
appropriate number of bins to build an effective model.

Furthermore, the total number of bins is determined by the number of
bins along the x-axis and y-axis (see \textbf{?@eq-equation13}). To
calculate the effective total number of bins, the minimum and
approximate maximum number of bins along the x-axis are used as
candidate values. The minimum number of bins along the x-axis is \(1\).
The maximum number of bins along the x-axis can be estimated by taking
the square root of the number of NLDR data points. To investigate the
impact of the number of bins on MSE, the analysis is focused on the MSE
for different total numbers of bins. This range encompasses the total
number of bins calculated for the minimum and maximum values of the
number of bins along the x-axis.

Typically, the MSE plot will have a steep slope at the beginning,
indicating that a smaller number of bins causes a larger amount of
error. Then, the slope will gradually decline or level off, indicating
that a higher number of bins generates a smaller error. This MSE plot is
useful in determining the effective number of bins required to construct
the 2D model. \textbf{?@fig-diagnosticpltScurve} (b) has created a graph
that shows how the Mean Squared Error (MSE) changes depending on the
number of bins assigned in the hexagonal grid for various NLDR
techniques applied to the S-curve data set. As per the rule of thumb
mentioned earlier, the effective number of bins for each NLDR method is
shown in the \textbf{?@tbl-msebinsnldr}. Furthermore,
\textbf{?@fig-modelScurve} shows the 2D models constructed for each NLDR
technique with the selected total number of bins and how the models look
in high-D with original data.

@@

The benchmark value for removing low-density hexagons has a range of
\(0\) and \(1\). When analyzing how these benchmark values affect model
performance, it's important to observe the change in Mean Squared Error
(MSE) as the benchmark value increases. The MSE exhibits a gradual
increase as the benchmark value progresses from \(0\) to \(1\).
Assessing this rate of increase is crucial. If the increment is not
substantial, the decision might lean towards retaining low-density
hexagons. According to \textbf{?@fig-diagnosticpltScurvelwd}, the change
in MSE for NLDR techniques is relatively small across different
benchmark values, except for the PHATE method.

Consequently, it is not necessary to remove low-density hexagons when
constructing the 2D model for the S-curve dataset in NLDR techniques,
except for PHATE. While there may be some fluctuations in MSE between
benchmark values of 0 to 1, the absence of significant changes suggests
that maintaining low-density hexagons is preferable. However, to choose
the best selection, it's recommended to examine the benchmark values
around the first local minimum in the MSE curve. However, the MSE for
PHATE fluctuates when benchmark values range between 0 and 1. Therefore,
it is necessary to explore the benchmark values surrounding the first
local minimum to determine an effective benchmark value. There is no
universal rule for selecting a process. As shown in
\textbf{?@fig-diagnosticpltScurvelwd}, the MSE values for PHATE
fluctuate, and the first local minimum is at \(0\). Therefore, nearby
benchmark values such as \(0.01\), \(0.04\), and \(0.06\) were explored
to identify the best value for preserving the data structure. In this
case, a benchmark value of \(0.06\) was chosen. The resulting 2D model
is shown in \textbf{?@fig-modelScurve} (c), while the video link
provides a visual representation of how the model appears in high-D.

@@

@@

Achieving a smooth representation in 2D space is crucial, and one factor
influencing this smoothness is the presence of long edges (see
\textbf{?@fig-modelScurvermlgimp}). These long edges occur when a line
connects points that are distant in the triangular mesh, impacting the
overall smoothness of the mesh.

@@

\hypertarget{starting-point-of-the-hexagonal-grid}{%
\subsubsection{Starting point of the hexagonal
grid}\label{starting-point-of-the-hexagonal-grid}}

According to \citet{Dan2023}, the hexagonal binning is done by
tessellating the \(xy\) plane over the set (range(\(x\)), range(\(y\)))
(see \textbf{?@fig-scurveshifthexgridsexp} (b)). In that case, bin
centroids are defined as shown in \textbf{?@fig-scurveshifthexgridsexp}
(a) with gray colour. Rather than sticking to the typical hexagonal
grid, introducing a meaningful shift in both the x and y directions
presents an opportunity for an improved 2D model. Therefore,
investigating this shift in the hexagonal grid is an important parameter
to consider.

As shown in \textbf{?@fig-scurveshifthexgrids}, the shifting influences
the distribution of points and number of non-empty bins, impacting the
resulting 2D model. According to
\textbf{?@fig-diagnosticpltScurvehexbins}, the 2D model with a total of
\(144\) bins applied to S-curve UMAP data does not require any shifting
because the lowest MSE occurs when no shift is introduced.

@@

In the 2D representations created by all NLDR techniques, as shown in
\textbf{?@fig-nldervis5Gau}, except for PHATE, there are five distinct
clusters. In tSNE, these five clusters are closely located to each other
(see \textbf{?@fig-nldervis5Gau} (a)). However, in PHATE, two clusters
are closely positioned, while the other three are more distant (see
\textbf{?@fig-nldervis5Gau} (c)). In PaCMAP, one cluster is at the
center, and the remaining four are positioned in four different
directions (see \textbf{?@fig-nldervis5Gau} (e)). In TriMAP, two
clusters are close, although not as close as in PHATE, and the other
three are well-separated (see \textbf{?@fig-nldervis5Gau} (d)). In UMAP,
all clusters are arranged in a parallel manner, with three in one line
and the other two in a separate line (see \textbf{?@fig-nldervis5Gau}
(b)).

Visualizing the models alongside the original high-D data provides
insights into how different techniques capture the underlying clustering
structure. The tSNE model exhibits five well-separated clusters,
effectively preserving both local and global structures (see video link
of \textbf{?@fig-modelfiveGau} (a)). UMAP, on the other hand, also
presents five distinct clusters but with a more flattened surface
appearance (see video link of \textbf{?@fig-modelfiveGau} (b)). The
PHATE model shows five separated clusters resembling triangles or
partial triangles but struggles to capture local structures (see video
link of \textbf{?@fig-modelfiveGau} (c)). TriMAP and PaCMAP both display
five well-separated clusters with flat surfaces, each capturing
within-cluster variation to varying extents. Despite the various 2D
representations, all NLDR techniques preserve the global structure (see
video link of \textbf{?@fig-modelfiveGau} (d), (e) respectively).
However, tSNE, effectively capture both local and global structures, as
indicated by lower AIC values in \textbf{?@fig-diagnosticpltGau} (a).

@@

\includegraphics{paper_files/figure-pdf/unnamed-chunk-55-1.pdf}

\includegraphics{paper_files/figure-pdf/unnamed-chunk-55-2.pdf}

\includegraphics{paper_files/figure-pdf/unnamed-chunk-55-3.pdf}

\includegraphics{paper_files/figure-pdf/unnamed-chunk-55-4.pdf}

@@

@@

\includegraphics{paper_files/figure-pdf/unnamed-chunk-61-1.pdf}

\includegraphics{paper_files/figure-pdf/unnamed-chunk-61-2.pdf}

\includegraphics{paper_files/figure-pdf/unnamed-chunk-61-3.pdf}

\includegraphics{paper_files/figure-pdf/unnamed-chunk-61-4.pdf}

@@

\includegraphics{paper_files/figure-pdf/unnamed-chunk-64-1.pdf}
